---
title: "Cyclistic Case Study"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Business Task

determine the difference between casual bike usage and annual member

# Stakeholders

-   executive team
-   director of marketing
-   marketing analytics team

# Data Source

The [data](https://divvy-tripdata.s3.amazonaws.com/index.html) has been
made available by Motivate International Inc. under this
[license](https://divvybikes.com/data-license-agreement).

The html page lists multiple data sets in .zip file, in order to
accomplish the business task, I am going to use the most recent data
which contains the 11 months trip data:
(`202301-divvy-tripdata.zip) - (202311-divvy-tripdata.zip)`.

With large size of data set, the spreadsheet is not ideal to finish the
task. Therefore, RStudio is preferred.

```{r load-library}
library(tidyverse)
library(skimr)  # summary stats
library(stringr)  # work with text
library(lubridate)  # extract date and time
library(ggplot2)  # visualization
library(forcats)  # recorder graph by value
```

# Load Data

There are 11 separated files indicate the trip condition of each month,
so the first step is to combine the data into a whole data set.

The structure of data is shown below. From the summary statistics, there
are

-   *7 categorical variables*

-   *4 numeric variables*

-   *2 datetime variables*

-   some variables contains bunch *of missing values*

    -   `end_lng`(ending longitude) and `end_lat`(ending latitude)
        contains 6751 missings

    -   about *15% to 17% of incomplete rate* in four categorical
        variables including
        `start_station_name, start_station_id, end_station_name, end_station_id`

```{r load-data}
tripdata_01 <- read_csv("202301-divvy-tripdata.csv")
tripdata_02 <- read_csv("202302-divvy-tripdata.csv")
tripdata_03 <- read_csv("202303-divvy-tripdata.csv")
tripdata_04 <- read_csv("202304-divvy-tripdata.csv")
tripdata_05 <- read_csv("202305-divvy-tripdata.csv")
tripdata_06 <- read_csv("202306-divvy-tripdata.csv")
tripdata_07 <- read_csv("202307-divvy-tripdata.csv")
tripdata_08 <- read_csv("202308-divvy-tripdata.csv")
tripdata_09 <- read_csv("202309-divvy-tripdata.csv")
tripdata_10 <- read_csv("202310-divvy-tripdata.csv")
tripdata_11 <- read_csv("202311-divvy-tripdata.csv")

tripdata <- 
  tripdata_01 %>% 
  rbind.data.frame(tripdata_02) %>% 
  rbind.data.frame(tripdata_03) %>% 
  rbind.data.frame(tripdata_04) %>% 
  rbind.data.frame(tripdata_05) %>% 
  rbind.data.frame(tripdata_06) %>% 
  rbind.data.frame(tripdata_07) %>% 
  rbind.data.frame(tripdata_08) %>% 
  rbind.data.frame(tripdata_09) %>% 
  rbind.data.frame(tripdata_10) %>% 
  rbind.data.frame(tripdata_11)

skim_without_charts(tripdata)
```

## Data Cleaning

The over view of column names are shown below

```{r name-of-columns}
colnames(tripdata)
```

I am going to do:

1.  remove duplicates
2.  remove records where end time is earlier than start time
3.  remove records with missing values

#### Remove duplicates

```{r remove_duplicates}
trip_no_dup <- 
  tripdata %>% 
  distinct()
```

#### Remove duration less than 0

```{r remove-duration-less-than-0}
trip_norm <- trip_no_dup %>% 
  mutate(duration_min = round(as.numeric((ended_at - started_at)/60), 2)) %>% # convert second to min
  mutate(day_of_week = weekdays(started_at)) %>% 
  filter(duration_min > 0)
```

After removing the 0-duration, I see that there are outliers in both
upper bound and lower bound for the `duration_min` variable. Therefore,
I go ahead and check the percentage where duration is less than 1 minute
and more than 24 hours to make sure they do not account for lots of data
points.

```{r stats-of-duration}
# summary of duration in minutes
trip_norm %>% 
  select(duration_min) %>% 
  summary()
```

```{r calculate-the-duration-boundary}
# calculate the ratio where duration is less than 1 minute
less_than_1 <- trip_norm %>% 
  filter(duration_min < 1) %>% 
  nrow() /
  nrow(trip_norm)

# calculate the ratio where people use over 24 hrs
greater_than_24 <- trip_norm %>% 
  filter(duration_min >= 24 * 60) %>% 
  nrow() / 
  nrow(trip_norm)

paste("the duration less than 1 is about ", round(less_than_1*100, 2), "% of the overall records")
paste("the duration less than 1 is about ", round(greater_than_24*100, 2), "% of the overall records")
```

After the calculation, and I see that the total duration which include
less than 1 minute and greater than 24 hours only account for no more
than 3% of data points. Therefore, these data is good to disappear.

In addition, I am interested in the time range of bike using. So, the
data has been added another feature which extracted from the
`started_at` and `ended_at`. The time has been splitted into 4 parts:
morning, afternoon, evening and nighttime.

```{r keep-duration-in-bound}
# define the time range for morning, afternoon, evening and nighttime
trip_norm <- trip_norm %>% 
  filter(duration_min >= 1 & duration_min < 24 * 60) %>% 
  mutate(started_date = as.Date(started_at),
         ended_date = as.Date(ended_at),
         started_time = paste(hour(started_at), minute(started_at), second(started_at), sep = ":"),
         ended_time = paste(hour(ended_at), minute(ended_at), second(ended_at), sep = ":"),
         time_range = ifelse(started_time >= "6:00:00" & started_time < "12:00:00", "morning(6am-12pm)",
                             ifelse(started_time >= "12:00:00" & started_time < "18:00:00", "afternoon(12pm-6pm)",
                                    ifelse(started_time >= "18:00:00" & started_time < "21:00:00", "evening(6pm-9pm)", "nighttime(9pm-6am)"))))
  
```

#### Remove missing value

```{r remove-missing-values}
trip_clean <- trip_norm %>% 
  filter(str_length(start_station_id) > 0) %>% 
  filter(str_length(end_station_id) > 0) %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, started_at, ended_at, started_time, ended_time, ride_id, start_station_id, end_station_id))  # remove longitude and latitude
```

#### Summary After Cleaning

```{r summary-after-cleaning}
skim_without_charts(trip_clean)

# save file locally
# write_csv(trip_clean, "/Users/phoebezou/Documents/google_certificate/DA-DS_case_study/cyclistic_case/trip_clean.csv")
```

# Analyze Data

For the annual member, there are only two bike types being used: classic
and electric

for the non-member user, there are three bike types being used: classic,
electric and docked

Meanwhile, the number of member user is about double amount than the
number of non-member user

#### Rideable type

```{r}
# the difference in rideable_type by different plans
trip_clean %>% 
  select(rideable_type, member_casual) %>% 
  count(member_casual, rideable_type) %>% 
  pivot_wider(names_from = rideable_type, values_from = n) %>% 
  replace(is.na(.), 0)

# the overall difference by different plans
trip_clean %>% 
  select(member_casual) %>% 
  count(member_casual) %>% 
  mutate(percentage = n / sum(n))
```

#### Investigate NA

Also, Base - 2132 W Hubbard is the only start station which member users
start riding a bike, that is why some aggregation results appear 'NA'

```{r}
# check why NA appear in some rows
trip_clean %>% 
  select(member_casual, start_station_name) %>% 
  filter(start_station_name == "Base - 2132 W Hubbard")

```

#### Start and end station

Here, I am going to check if there are different start station
preferences for both user type, and I replace NA with 0 because I find
out that some station only member start with them. In addition, member
users are more than non-member user in the most start station.

```{r}
# start station
pop_start_station <- 
  trip_clean %>% 
  select(start_station_name, member_casual) %>%   # get station name and membership status
  count(start_station_name, member_casual) %>%  # group and count by these value
  arrange(start_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by station name
pop_start_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_start_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_start_station %>%
  arrange(desc(casual)) %>%
  head(10)
```

```{r}
# end station
pop_end_station <- 
  trip_clean %>% 
  select(end_station_name, member_casual) %>%   # get station name and membership status
  count(end_station_name, member_casual) %>%  # group and count by these value
  arrange(end_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by stai=tion name
pop_end_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_end_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_end_station %>%
  arrange(desc(casual)) %>%
  head(10)
```

Moreover, I found out that the morning slot has not records which may
indicate that member and non-member users do not like to ride bikes in
the morning, or maybe they tend to choose other vihecle to commute.

```{r}
# double check the morning slot
trip_clean %>% 
  filter(time_range == "morning(6am-12pm)")
```

# Visualization

#### User Type Ratio

After all the cleaning, the overall ratio of user type indicates that
member users are a lot more than non-member users. Since the data set
does not contain the amount that users pay for, therefore, there is no
way for me to understand how the price difference between the two.

However, it is better to retain current member and obtain potential
member to maximize the profit for a long run.

```{r user-type-ratio}
trip_clean %>% 
  count(member_casual) %>% 
  mutate(membership_percentage = round(n/sum(n)*100,2)) %>% 
  ggplot(aes(x = member_casual, y = membership_percentage, label = membership_percentage)) +
  geom_col() +
  geom_text(vjust = -0.5) +
  theme_minimal() + 
  labs(title = "Ratio of user types",
       x = "user types",
       y = "percentage") 
  
```

#### Top 5 start/end station

the top 5 station is determined by the total number of users. Compare
both plots, the popular start station and end station are the same. In
addition, it is interesting that *Streeter Dr & Grand Ave* has large
amount of casual users, and member users are more than non-member user
in *Clark St & Elm St*. However, there are no corresponding geographic
datasets to see the location and distance, otherwise, I may be able to
see why such phenomenon.

```{r top-5-start-station}
pop_start_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(start_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(start_station_name, value)), y = value)) + # reorder the name of station by the value
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 start station names with different user types",
       x = "start station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

```{r top-5-end-station}
pop_end_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(end_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(end_station_name, value)), y = value)) + 
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 end station names with different user types",
       x = "end station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```
#### Days

  From the below statistics, the usage of bike seems roughly equally distributed. However, as the plot shows that there are more member users on Monday through Thursday, and it turns opposite from Friday to Sunday. 
  
  The trend to use bikes in days between two types of users are opposite as well. Member users seem to use bike in weekdays and the amount of usage decrease on weekends. Moreover, non-member users are more likely to use bike when the day approach to weekends and Saturday is the peak.
  
```{r differentiate-by-days}
# differentiate by user type
trip_clean %>% 
  group_by(member_casual) %>% 
  count(day_of_week) %>% 
  arrange(desc(n)) %>% 
  mutate(n = n/sum(n),
         day_of_week = fct_reorder(day_of_week, n)) %>% 
  ggplot(aes(x = day_of_week, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "bike usage in days differetiated by user types",
       x = "days",
       y = "percentage") +
  theme_minimal() +
  scale_fill_discrete(name = "user type") # change legend name
 
# bike use in days in general
trip_clean %>% 
  count(day_of_week) %>% 
  mutate(n = n/sum(n)) %>% 
  arrange(desc(n))
```
#### Time range in a day

  The general plot shows that most of users like to use bikes around 12pm-6pm. During the time, as the user type plot shows that casual users are more than member users, and member users are more than non-member users during nighttime. But it cannot conclude, without more information, that member users like to use bike in nighttime.
```{r differentiate-by-time-range}
# time range in general
trip_clean %>% 
  count(time_range) %>% 
  mutate(n = n/sum(n),
         time_range = fct_reorder(time_range, desc(n))) %>% 
  ggplot(aes(x = time_range, y = n)) + 
  geom_col() + 
  theme_minimal() + 
  labs(x = "time range",
       y = "numer of bike use in percentage",
       title = "bike usage in different time range in general")

# time range by different user types
trip_clean %>% 
  group_by(member_casual) %>% 
  count(time_range) %>% 
  mutate(n = n/sum(n),
         time_range = fct_reorder(time_range, desc(n))) %>% 
  ggplot(aes(x = time_range, y = n, fill = member_casual)) + 
  geom_col(position = "dodge") + 
  theme_minimal() + 
  labs(x = "time range",
       y = "numer of bike use in percentage",
       title = "bike usage in different time range by user type") +
  scale_fill_discrete(name = "user type") # change legend name
```
#### Month

  the plot below shows the difference in month that the condition of bike usage for both type of users. I would say that the individual blue and red bar are roughly bell curves, the blue(member) spreads wider, and the red(casual) is more center. Both type of users increase from the 4th month and reach the peak(the 7th month) then decrease.
```{r differentiate-in-month}
trip_clean %>% 
  mutate(start_month = month(started_date)) %>% 
  select(start_month, member_casual) %>% 
  count(member_casual, start_month) %>% 
  group_by(member_casual) %>% 
  mutate(n = n / sum(n)) %>% 
  ggplot(aes(start_month, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_discrete(name = "user type") +
  labs(x = "start month",
       y = "bike usage in percentage",
       title = "popular month to use bike by different user types") +
  theme_minimal()
```
