---
title: "Cyclistic Case Study"
output: html_notebook
---
# Business Task
  
  determine the difference between casual bike usage and annual member
  
# Stakeholders
  * executive team
  * director of marketing
  * marketing analytics team
  
# Data Source

  The [data](https://divvy-tripdata.s3.amazonaws.com/index.html) has been made available by Motivate International Inc. under this [license](https://divvybikes.com/data-license-agreement). 
  
  The html page lists multiple datasets in .zip file, in order to accomplish the business task, I am going to use the most recent data which contains the 11 months trip data, downloadable from `202301-divvy-tripdata.zip - 202311-divvy-tripdata.zip`. 
  
  With large size of dataset, the spreadsheet is not ideal to finish the task. Therefore, RStudio is preferred.
```{r load-library}
library(tidyverse)
library(skimr)
library(stringr)
library(lubridate)  # extract date and time
library(ggplot2)  # visualization
library(forcats)  # recorder bar-graph by value
```

# Load Data

  There are 11 separated files indicate the trip condition of each month, so the first step is to combine the data into a whole dataset.
```{r load-data}
tripdata_01 <- read_csv("202301-divvy-tripdata.csv")
tripdata_02 <- read_csv("202302-divvy-tripdata.csv")
tripdata_03 <- read_csv("202303-divvy-tripdata.csv")
tripdata_04 <- read_csv("202304-divvy-tripdata.csv")
tripdata_05 <- read_csv("202305-divvy-tripdata.csv")
tripdata_06 <- read_csv("202306-divvy-tripdata.csv")
tripdata_07 <- read_csv("202307-divvy-tripdata.csv")
tripdata_08 <- read_csv("202308-divvy-tripdata.csv")
tripdata_09 <- read_csv("202309-divvy-tripdata.csv")
tripdata_10 <- read_csv("202310-divvy-tripdata.csv")
tripdata_11 <- read_csv("202311-divvy-tripdata.csv")

tripdata <- 
  tripdata_01 %>% 
  rbind.data.frame(tripdata_02) %>% 
  rbind.data.frame(tripdata_03) %>% 
  rbind.data.frame(tripdata_04) %>% 
  rbind.data.frame(tripdata_05) %>% 
  rbind.data.frame(tripdata_06) %>% 
  rbind.data.frame(tripdata_07) %>% 
  rbind.data.frame(tripdata_08) %>% 
  rbind.data.frame(tripdata_09) %>% 
  rbind.data.frame(tripdata_10) %>% 
  rbind.data.frame(tripdata_11)
```

# Summary of Data

The structure of data is shown below. From the summary statistics, there are 
  * *7 categorical variables, 4 numeric variables, and 2 datetime variables*
  * both *6751 missing values* in `end_lng`(ending longitute) and `end_lat`(ending latitute)
  * about *15% to 17% of incomplete rate* in four categorical variables including `start_station_name, start_station_id, end_station_name, end_station_id`
```{r}
skim_without_charts(tripdata)
```

## Data Cleaning 

  The name of columns are listed below. We are going to do:
  1. remove duplicates
  2. remove records where end time is earlier than start time
  3. remove records with missing values
  
```{r name-of-columns}
colnames(tripdata)
```
#### Remove Duplicates
```{r}
trip_no_dup <- 
  tripdata %>% 
  distinct()
```
#### Remove Duration Less than 0
```{r calculate-ride-duration}
trip_norm <- trip_no_dup %>% 
  mutate(duration_min = round(as.numeric((ended_at - started_at)/60), 2)) %>% 
  mutate(day_of_week = weekdays(started_at)) %>% 
  filter(duration_min > 0)
```
```{r}
# define the time range for morning, afternoon, evening and nighttime
trip_norm <- trip_norm %>% 
  mutate(started_date = as.Date(started_at),
         ended_date = as.Date(ended_at),
         started_time = paste(hour(started_at), minute(started_at), second(started_at), sep = ":"),
         ended_time = paste(hour(ended_at), minute(ended_at), second(ended_at), sep = ":"),
         time_range = ifelse(started_time >= "6:00:00" & started_time < "12:00:00", "morning(6am-12pm)",
                             ifelse(started_time >= "12:00:00" & started_time < "18:00:00", "afternoon(12pm-6pm)",
                                    ifelse(started_time >= "18:00:00" & started_time < "21:00:00", "evening(6pm-9pm)", "nighttime(9pm-6am)"))))
  
```

#### Remove missing values
```{r}
trip_clean <- trip_norm %>% 
  filter(str_length(start_station_id) > 0) %>% 
  filter(str_length(end_station_id) > 0) %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, started_at, ended_at, started_time, ended_time))  # remove longitude and latitude
```
#### Summary After Cleaning
```{r}
skim_without_charts(trip_clean)

write_csv(trip_clean, "/Users/phoebezou/Documents/google_certificate/DA-DS_case_study/cyclistic_case/2023_tripClean.csv")
```

# Analyze Data

```{r}
colnames(trip_clean)  # double check the column name
```
```{r}
# summary of duration in minutes
trip_clean %>% 
  select(duration_min) %>% 
  summary()
```


```{r}
# the difference in rideable_type by different plans
trip_clean %>% 
  select(rideable_type, member_casual) %>% 
  count(member_casual, rideable_type) %>% 
  pivot_wider(names_from = rideable_type, values_from = n) %>% 
  replace(is.na(.), 0)

# the overall difference by different plans
trip_clean %>% 
  select(member_casual) %>% 
  count(member_casual)
```
```{r}
# check why NA appear in some rows
trip_clean %>% 
  select(member_casual, start_station_name) %>% 
  filter(start_station_name == "Base - 2132 W Hubbard")
```

```{r}
# start station
pop_start_station <- 
  trip_clean %>% 
  select(start_station_name, member_casual) %>%   # get station nane and membership status
  count(start_station_name, member_casual) %>%  # group and count by these value
  arrange(start_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by station name
pop_start_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_start_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_start_station %>%
  arrange(desc(casual)) %>%
  head(10)
```
```{r}
# end station
pop_end_station <- 
  trip_clean %>% 
  select(end_station_name, member_casual) %>%   # get station nane and membership status
  count(end_station_name, member_casual) %>%  # group and count by these value
  arrange(end_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by stai=tion name
pop_end_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_end_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_end_station %>%
  arrange(desc(casual)) %>%
  head(10)
```

```{r}
# started date

# check if a ride is over a day period
trip_clean %>%
  filter(duration_min >= 24 * 60)

# check if a ride is over a month
trip_clean %>%
  filter(duration_min >= 24 * 60 * 30)

pop_started_month <- trip_clean %>%
  select(member_casual, started_date) %>%
  mutate(start_month = month(started_date)) %>%
  select(-started_date) %>% 
  count(member_casual, start_month)
```

```{r}
# time range
pop_time_range <- trip_clean %>% 
  select(member_casual, time_range) %>% 
  count(member_casual, time_range)

# double check the morning slot
trip_clean %>% 
  filter(time_range == "morning(6am-12pm)")
```

# Visualization
```{r user-type-ratio}
trip_clean %>% 
  count(member_casual) %>% 
  mutate(membership_percentage = round(n/sum(n)*100,2)) %>% 
  ggplot(aes(x = member_casual, y = membership_percentage, label = membership_percentage)) +
  geom_col() +
  geom_text(vjust = -0.5) +
  theme_minimal() + 
  labs(title = "Ratio of user types",
       x = "user types",
       y = "percentage") 
  
```



```{r top-5-start-station}
pop_start_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(start_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(start_station_name, value)), y = value)) + # reorder the name of station by the value
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 start station names with different user types",
       x = "start station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```
```{r top-5-end-station}
pop_end_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(end_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(end_station_name, value)), y = value)) + 
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 end station names with different user types",
       x = "end station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

```{r bike-use-over-a-day}
# total 130 records of bike usages over 24 hrs
trip_clean %>% 
  filter(duration_min >= 24*60) %>% 
  nrow()

# the ratio of user types for these records
trip_clean %>% 
  filter(duration_min >= 24*60) %>% 
  count(member_casual) %>% 
  ggplot(aes(x = member_casual, y = n)) +
  geom_col() +
  labs(x = "user type",
       y = "count",
       title = "bike usage over 24 hours") +
  theme_minimal()
```
```{r}
trip_clean %>% 
  filter(duration_min < 24 * 60 & duration_min >= 1) %>% 
  ggplot(aes(x = duration_min)) + 
  geom_histogram()
```
```{r}
# calculate the ratio where duration is less than 1 minute
trip_clean %>% 
  filter(duration_min < 1) %>% 
  nrow() /
  nrow(trip_clean)

# calculate the ratio where people use over 24 hrs
trip_clean %>% 
  filter(duration_min >= 24 * 60) %>% 
  nrow() / 
  nrow(trip_clean)
```

