---
title: "Cyclistic Case Study"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Business Task

determine the difference between casual bike usage and annual member

# Stakeholders

-   executive team
-   director of marketing
-   marketing analytics team

# Data Source

The [data](https://divvy-tripdata.s3.amazonaws.com/index.html) has been
made available by Motivate International Inc. under this
[license](https://divvybikes.com/data-license-agreement).

The html page lists multiple data sets in .zip file, in order to
accomplish the business task, I am going to use the most recent data
which contains the 11 months trip data:
(`202301-divvy-tripdata.zip) - (202311-divvy-tripdata.zip)`.

With large size of data set, the spreadsheet is not ideal to finish the
task. Therefore, RStudio is preferred.

```{r load-library}
library(tidyverse)
library(skimr)  # summary stats
library(stringr)  # work with text
library(lubridate)  # extract date and time
library(ggplot2)  # visualization
library(forcats)  # recorder graph by value
```

# Load Data

There are 11 separated files indicate the trip condition of each month,
so the first step is to combine the data into a whole data set.

The structure of data is shown below. From the summary statistics, there
are

-   *7 categorical variables*

-   *4 numeric variables*

-   *2 datetime variables*

-   some variables contains bunch *of missing values*

    -   `end_lng`(ending longitude) and `end_lat`(ending latitude)
        contains 6751 missings

    -   about *15% to 17% of incomplete rate* in four categorical
        variables including
        `start_station_name, start_station_id, end_station_name, end_station_id`

```{r load-data}
tripdata_01 <- read_csv("202301-divvy-tripdata.csv")
tripdata_02 <- read_csv("202302-divvy-tripdata.csv")
tripdata_03 <- read_csv("202303-divvy-tripdata.csv")
tripdata_04 <- read_csv("202304-divvy-tripdata.csv")
tripdata_05 <- read_csv("202305-divvy-tripdata.csv")
tripdata_06 <- read_csv("202306-divvy-tripdata.csv")
tripdata_07 <- read_csv("202307-divvy-tripdata.csv")
tripdata_08 <- read_csv("202308-divvy-tripdata.csv")
tripdata_09 <- read_csv("202309-divvy-tripdata.csv")
tripdata_10 <- read_csv("202310-divvy-tripdata.csv")
tripdata_11 <- read_csv("202311-divvy-tripdata.csv")

tripdata <- 
  tripdata_01 %>% 
  rbind.data.frame(tripdata_02) %>% 
  rbind.data.frame(tripdata_03) %>% 
  rbind.data.frame(tripdata_04) %>% 
  rbind.data.frame(tripdata_05) %>% 
  rbind.data.frame(tripdata_06) %>% 
  rbind.data.frame(tripdata_07) %>% 
  rbind.data.frame(tripdata_08) %>% 
  rbind.data.frame(tripdata_09) %>% 
  rbind.data.frame(tripdata_10) %>% 
  rbind.data.frame(tripdata_11)

skim_without_charts(tripdata)
```

## Data Cleaning

The over view of column names are shown below

```{r name-of-columns}
colnames(tripdata)
```

I am going to do:

1.  remove duplicates
2.  remove records where end time is earlier than start time
3.  remove records with missing values

#### Remove duplicates

```{r remove_duplicates}
trip_no_dup <- 
  tripdata %>% 
  distinct()
```

#### Remove duration less than 0

```{r remove-duration-less-than-0}
trip_norm <- trip_no_dup %>% 
  mutate(duration_min = round(as.numeric((ended_at - started_at)/60), 2)) %>% # convert second to min
  mutate(day_of_week = weekdays(started_at)) %>% 
  filter(duration_min > 0)
```

After removing the 0-duration, I see that there are outliers in both
upper bound and lower bound for the `duration_min` variable. Therefore,
I go ahead and check the percentage where duration is less than 1 minute
and more than 24 hours to make sure they do not account for lots of data
points.

```{r stats-of-duration}
# summary of duration in minutes
trip_norm %>% 
  select(duration_min) %>% 
  summary()
```

```{r calculate-the-duration-boundary}
# calculate the ratio where duration is less than 1 minute
less_than_1 <- trip_norm %>% 
  filter(duration_min < 1) %>% 
  nrow() /
  nrow(trip_norm)

# calculate the ratio where people use over 24 hrs
greater_than_24 <- trip_norm %>% 
  filter(duration_min >= 24 * 60) %>% 
  nrow() / 
  nrow(trip_norm)

paste("the duration less than 1 is about ", round(less_than_1*100, 2), "% of the overall records")
paste("the duration less than 1 is about ", round(greater_than_24*100, 2), "% of the overall records")
```

After the calculation, and I see that the total duration which include
less than 1 minute and greater than 24 hours only account for no more
than 3% of data points. Therefore, these data is good to disappear.

In addition, I am interested in the time range of bike using. So, the
data has been added another feature which extracted from the
`started_at` and `ended_at`

```{r keep-duration-in-bound}
# define the time range for morning, afternoon, evening and nighttime
trip_norm <- trip_norm %>% 
  filter(duration_min >= 1 & duration_min < 24 * 60) %>% 
  mutate(started_date = as.Date(started_at),
         ended_date = as.Date(ended_at),
         started_time = paste(hour(started_at), minute(started_at), second(started_at), sep = ":"),
         ended_time = paste(hour(ended_at), minute(ended_at), second(ended_at), sep = ":"),
         time_range = ifelse(started_time >= "6:00:00" & started_time < "12:00:00", "morning(6am-12pm)",
                             ifelse(started_time >= "12:00:00" & started_time < "18:00:00", "afternoon(12pm-6pm)",
                                    ifelse(started_time >= "18:00:00" & started_time < "21:00:00", "evening(6pm-9pm)", "nighttime(9pm-6am)"))))
  
```

#### Remove missing value

```{r remove-missing-values}
trip_clean <- trip_norm %>% 
  filter(str_length(start_station_id) > 0) %>% 
  filter(str_length(end_station_id) > 0) %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng, started_at, ended_at, started_time, ended_time))  # remove longitude and latitude
```

#### Summary After Cleaning

```{r summary-after-cleaning}
skim_without_charts(trip_clean)

# save file locally
# write_csv(trip_clean, "/Users/phoebezou/Documents/google_certificate/DA-DS_case_study/cyclistic_case/trip_clean.csv")
```

# Analyze Data

For the annual member, there are only two bike types being used: classic
and electric

for the non-member user, there are three bike types being used: classic,
electric and docked

Meanwhile, the number of member user is about double amount than the
number of non-member user

```{r}
# the difference in rideable_type by different plans
trip_clean %>% 
  select(rideable_type, member_casual) %>% 
  count(member_casual, rideable_type) %>% 
  pivot_wider(names_from = rideable_type, values_from = n) %>% 
  replace(is.na(.), 0)

# the overall difference by different plans
trip_clean %>% 
  select(member_casual) %>% 
  count(member_casual) %>% 
  mutate(percentage = n / sum(n))
```

```{r}
# check why NA appear in some rows
trip_clean %>% 
  select(member_casual, start_station_name) %>% 
  filter(start_station_name == "Base - 2132 W Hubbard")
```

```{r}
# start station
pop_start_station <- 
  trip_clean %>% 
  select(start_station_name, member_casual) %>%   # get station name and membership status
  count(start_station_name, member_casual) %>%  # group and count by these value
  arrange(start_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by station name
pop_start_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_start_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_start_station %>%
  arrange(desc(casual)) %>%
  head(10)
```

```{r}
# end station
pop_end_station <- 
  trip_clean %>% 
  select(end_station_name, member_casual) %>%   # get station nane and membership status
  count(end_station_name, member_casual) %>%  # group and count by these value
  arrange(end_station_name) %>% 
  pivot_wider(names_from = member_casual, values_from = n) %>%  # transform table
  as.data.frame() %>%
  replace(is.na(.),0) %>% # fill na w. 0
  mutate(member_greater_than_casual = member - casual > 0,
         total_user = casual + member)  # get stats

# overall no. of member stats vs no. of casual by stai=tion name
pop_end_station %>%
  count(member_greater_than_casual)

# top 10 start station for member
pop_end_station %>%
  arrange(desc(member)) %>%
  head(10)

# top 10 start station for casual
pop_end_station %>%
  arrange(desc(casual)) %>%
  head(10)
```

```{r}
# double check the morning slot
trip_clean %>% 
  filter(time_range == "morning(6am-12pm)")
```

# Visualization

```{r user-type-ratio}
trip_clean %>% 
  count(member_casual) %>% 
  mutate(membership_percentage = round(n/sum(n)*100,2)) %>% 
  ggplot(aes(x = member_casual, y = membership_percentage, label = membership_percentage)) +
  geom_col() +
  geom_text(vjust = -0.5) +
  theme_minimal() + 
  labs(title = "Ratio of user types",
       x = "user types",
       y = "percentage") 
  
```

```{r top-5-start-station}
pop_start_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(start_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(start_station_name, value)), y = value)) + # reorder the name of station by the value
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 start station names with different user types",
       x = "start station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

```{r top-5-end-station}
pop_end_station %>% 
  pivot_longer(cols = c(casual, member), names_to = "user_types") %>% 
  group_by(end_station_name) %>% 
  arrange(desc(total_user)) %>% 
  head(10) %>% 
  ggplot(aes(x = fct_rev(fct_reorder(end_station_name, value)), y = value)) + 
  geom_col(aes(fill = user_types), position = "dodge") +
  labs(title = "Top 5 end station names with different user types",
       x = "end station",
       y = "number of users") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
```

```{r differentiate-by-days}
# differentiate by user type
trip_clean %>% 
  group_by(member_casual) %>% 
  count(day_of_week) %>% 
  arrange(desc(n)) %>% 
  mutate(n = n/sum(n),
         day_of_week = fct_reorder(day_of_week, n)) %>% 
  ggplot(aes(x = day_of_week, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "bike usage in days differetiated by user types",
       x = "days",
       y = "percentage") +
  theme_minimal() +
  scale_fill_discrete(name = "user type") # change legend name
 
# bike use in days in general
trip_clean %>% 
  filter(duration_min >= 1 & duration_min < 24 * 60) %>% 
  count(day_of_week) %>% 
  mutate(n = n/sum(n)) %>% 
  arrange(desc(n))
    
```

```{r differentiate-by-time-range}
# time range in general
trip_clean %>% 
  count(time_range) %>% 
  mutate(n = n/sum(n),
         time_range = fct_reorder(time_range, desc(n))) %>% 
  ggplot(aes(x = time_range, y = n)) + 
  geom_col() + 
  theme_minimal() + 
  labs(x = "time range",
       y = "numer of bike use in percentage",
       title = "bike usage in different time range in general")

# time range by different user types
trip_clean %>% 
  group_by(member_casual) %>% 
  count(time_range) %>% 
  mutate(n = n/sum(n),
         time_range = fct_reorder(time_range, desc(n))) %>% 
  ggplot(aes(x = time_range, y = n, fill = member_casual)) + 
  geom_col(position = "dodge") + 
  theme_minimal() + 
  labs(x = "time range",
       y = "numer of bike use in percentage",
       title = "bike usage in different time range in general") +
  scale_fill_discrete(name = "user type") # change legend name
```

```{r differentiate-in-month}
trip_clean %>% 
  mutate(start_month = month(started_date)) %>% 
  select(start_month, member_casual) %>% 
  count(member_casual, start_month) %>% 
  group_by(member_casual) %>% 
  mutate(n = n / sum(n)) %>% 
  ggplot(aes(start_month, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_discrete(name = "user type") +
  labs(x = "start month",
       y = "bike usage in percentage",
       title = "popular month to use bike by different user types") +
  theme_minimal()
```
